# Results

The main focus of our analysis will examine Robintrack's users stock holding of the S&P500 companies between August 2018 - August 2020. We will explore how price, market cap, industry and founding year affect user's holding.

## Average ownership per year by industry

```{r}
library(GGally)
library(ggplot2)
library(palmerpenguins)

data <-read.csv('docs/data/results/alluvial.csv')
```

The first transformed dataframe the average user's stock holding per industry for 2018, 2019 and 2020. 

In order explore the shifts in the most popular industries between 2018-2020, we have created a Clevland plot for which we have sorted the y axis in descending order for the users holding on 2020.

```{r qplot51, fig.width=10, fig.height=6,message=FALSE}
library(tidyverse)
data$Date<-as.factor(data[,'Date'])
# create the plot
ggplot(data, aes(x = users_holding, y =fct_reorder2(Symbol, Date == '2020', 
                                 users_holding, .desc=FALSE),color = Date)) +
    geom_point() +
    xlab("\nAverage users holding") +
    ylab("Industry\n") +
    ggtitle("Average yearly Stock Holding by industry\nAugust 2018 - August 2020") 
```

Some insights that we can extract from the above plot are:

* We can see a jump in users holding for across all industries between 2018/2019 and 2020. A possible explanation is that retail trading has taken off in 2020 amid the coronavirus downturn that many young traders saw as an entry point into the world of investing.

* This jump is especially evident in the following industries;`Comunication Services`, `Consumer Discretionary`,`Industrials` and `Energy`.

* The largest jump in user's holding between 2019 and 2020 was in the `Consumer Discretionary` industry. A possible explanation could be the increased interest of the Amazon (AMZN) stock during the pandemic.

* Even though  `Information Technology` industry had the second highest average user's holding in 2018 and 2019, in 2020 it dropped to 4th place.

* Similarly in 2018 & 2019 `Energy` industry was less attractive than `Consumer Staples`. However, in 2020 `Energy` attracted more stock holders. ``

* Another evident trend is that `Utilities`, `Materials` and `Real Estate` has similar average user's holding between 2018 and 2019. On the other hand the rest of the industried have experienced and increase in user's holding between these two years.

* Moreover, `Consumer Staples` and  `Industrials` industries have almost the same stock holding between 2018 and 2019. That trend was broken in 2020 when `Industrials` almost tripled in user's holding compared to `Consumer Staples`.


## Exploring the relationship of founding date vs popularity level 

As mentioned in the data transformation section, we measure a stock's popularity by averaging amount of user's holding acrross 2018-2020. We assign each company to a popularilty group by using KMeans clustering(k=3) for the avergage user's holding over those 3 years. The three popularity categories are; `not_popular`(n=473), `normal`(n=16) and `very_popular`(n=5).  

Moreover, as mentioned in the transformation section, we have also split the companies in three fouding dates ranges. We again used KMeans clustering(k=3) in order to group the S&P500 companies in three age groups for which we recorded the minimum and maximum founding rate in each group as their label. More specifically the three founding ranges are;`1784-1892`(n=247),`1894-1952`(n=400), `1954-2019`(n=798).

```{r qplot56, fig.width=6, fig.height=6,message=FALSE}
library(vcd)
mosaic <-read.csv('docs/data/results/mosaic.csv')
mosaic$Founded_range<-factor(mosaic$Founded_range, levels=c("1784-1892","1894-1952","1954-2019"))
mosaic$popularity<-factor(mosaic$popularity, levels=c("not_popular","normal","very_popular"))

colors = c("grey95", "aliceblue", "cornflowerblue")
vcd::mosaic(Founded_range~popularity, direction = c("v","h") ,mosaic,
            highlighting_fill = colors, gp_varnames = gpar(fontsize = 10, fontface = 1),gp_labels = gpar(fontsize = 8))
```

In the above mosaic plot we can see that there is a correlation between the age of a company and its popularity. More specifically:

* We can see that for the less popular companies there is an equal amount of observations across all three founding ranges.

* When comparing the normal and the less popular companies we can see an increase in popularity as the founding date becomes more recent.

* Moreover, even though the oldest copanies are the ones that appear the least across the very popular companies, we can see that the medium founding range companies (`1894-1952`) are the the most across the `very_popular`.

* We would have been able to identify a positive correlation between how recently a company was founded and how popular the company is amongst Robinhood users. However this trend is broken in the `very_popular` companies.

A possible explanation of the partial positive correlation between founding year and popularity could due to the fact that Robinhood’s average customer is young^[According to [this](https://www.nytimes.com/2020/07/08/technology/robinhood-risky-trading.html) article] and therefore is more aware of the more recently founded companies.

## A deeper look into the `very_popular` companies

```{r qplot57, fig.width=20, fig.height=10,message=FALSE}
# Libraries
library(ggplot2)
library(dplyr)
library("tidyverse")
library("ggridges")
library(scales)
library(lubridate)
ridge <- read.csv('docs/data/results/ridge.csv')
very_popular_ridge <- subset(ridge, ridge$popularity == 'very_popular')

very_popular_ridge$Date <- as.Date(very_popular_ridge$Date)
# Plot
very_popular_ridge %>%
  ggplot( aes(x=Date, y=users_holding, group=Symbol, color=Symbol)) +
  geom_line() +  
  scale_x_date(date_labels = "%b-%Y", breaks = "2 month") + 
  ggtitle("Daily stock user's holding for the most popular S&P500 companies\nAugust 2018 - August 2020") +
  theme(plot.title=element_text(size=22),
          axis.text=element_text(size=19),
         axis.title=element_text(size=18),
         legend.title = element_text( size = 16),
        legend.text = element_text(size = 14))
```


```{r qplot59,  fig.width=20, fig.height=10, message=FALSE}
very_popular_ridge %>%
  ggplot( aes(x=Date, y=Adj.Close, group=Symbol, color=Symbol)) +
  geom_line() +
  scale_x_date(date_labels = "%b-%Y", breaks = "2 month") +
  ggtitle("Daily stock price for the most popular S&P500 companies\nAugust 2018 - August 2020") +
  theme( plot.title=element_text(size=22),
          axis.text=element_text(size=19),
         axis.title=element_text(size=18),
         legend.title = element_text( size = 16),
        legend.text = element_text(size = 14))
```


```{r qplot58, fig.width=20, fig.height=11,message=FALSE}
very_popular_ridge %>%
  ggplot( aes(x=Date, y=Mcap, group=Symbol, color=Symbol)) +
  geom_line() +  
  scale_x_date(date_labels = "%b-%Y", breaks = "2 month") + 
  ggtitle("Daily Market Cap for the most popular S&P500 companies\nAugust 2018 - August 2020")  +
  theme( plot.title=element_text(size=22),
          axis.text=element_text(size=19),
         axis.title=element_text(size=18),
         legend.title = element_text( size = 16),
        legend.text = element_text(size = 14))
```


```{r qplot60, fig.width=10, fig.height=6, message=FALSE}

very_popular_ridge %>%
  ggplot( aes(x=Date, y=vix_Adj_close)) +
  geom_line() + 
  ggtitle("Daily VIX\nAugust 2018 - August 2020") +
  scale_x_date(date_labels = "%b-%Y", breaks = "2 month") + 
  theme( plot.title=element_text(size=11),
          axis.text=element_text(size=9),
         axis.title=element_text(size=8),
         legend.title = element_text( size = 6),
        legend.text = element_text(size = 5))
```
